---
title: "Overall Notes"
author: "Allyson Cameron"
date: "2022-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# some stuff you will never need
# install.packages("gitcreds")
# gitcreds::gitcreds_set()
library(rethinking)
library(tidybayes.rethinking)
library(ggplot2)
library(dplyr)
library(tidybayes)
library(tidyr)
library(tidyverse)
# install.packages("lme4")
library(lme4)
```


# Chapter 3

Use samples to summarize and simulate model output

```{r}

# here is our grid approximation of the water data.

p_grid <- seq(from = 0, to = 1, length.out = 1000)
prob_p <- rep(1, 1000)
prob_data <- dbinom(6, size = 9, prob = p_grid)
# 6 /9 is our likelihood of water ^
posterior <- prob_data * prob_p
posterior <- posterior / sum(posterior)

# now let's see sampling
```

I think this is how you go about doing a prior predictive simulation using quap

first our model: 

$$
\begin{aligned}
weight^*_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta \: height^*_i \\ 
\alpha &= 0 \\ 
\beta &\sim \text{Lognormal}(0,1) \\
\sigma &\sim \text{Uniform}(0,30) 
\end{aligned}
$$


```{r}
# how many fake lines?
nsims <- 12
# simulate some lines (and sigmas)
set.seed(0929)
sims2 <- tibble(
  id = 1:nsims,
  a = rep(0, nsims),
  b = rlnorm(nsims, meanlog = 0, sdlog = 1),
  sigma = runif(nsims, min = 0, max = 30)
)
# here are the fake lines (and sigmas)
head(sims2, nsims)
```

We can plot these lines.

```{r}
ggplot(sims2) +
  geom_abline(
    aes(
      color = factor(id),
      slope = b,
      intercept = a
    ),
    size = 1.5,
    alpha = .7
  ) +
  xlim(-20, 20) +
  ylim(-20, 20) +
  labs(
    x = "height",
    y = "weight"
  ) +
  theme(legend.position = "none") +
  theme(aspect.ratio = 1)
```

# Chapter 4

Posterior distribution of a linear model plotted against data with uncertainty = posterior predictive check.

```{r}
# 4.42 load data again, since it's a long way back
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[d$age >= 18, ]
d2 <- d2 |>
  mutate(
    height_z = standardize(height),
    weight_z = standardize(weight)
  )

# define the average weight,x-bar
xbar <- mean(d2$weight)
# fit model

m4.3 <- quap(
  alist(
    weight_z ~ dnorm(mu, sigma),
    mu <- a + b * (height_z),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ),
  data = d2
)

draws <- tidy_draws(m4.3, 100)

# plot OG data and predcited lines

p <- ggplot(draws) +
  geom_abline(
    aes(
      slope = b,
      intercept = a
    ),
    size = .2,
    alpha = .2
  ) +
  geom_point(data = d2, mapping = aes(x = height_z, y = weight_z), alpha = .2)

p
# adding uncertainty

# predicted draws grabs same values as before, but also with sigma attached
ppsims <- predicted_draws(m4.3,
  newdata = d2,
  draws = 1000
)
head(ppsims, 20)
View(ppsims)

# now we are using that stuff to make our boundaries for our plot
ppsims <- ppsims |>
  group_by(.row) |>
  mutate(
    lo_bound = HPDI(.prediction)[1],
    up_bound = HPDI(.prediction)[2]
  )

p + geom_ribbon(
  data = ppsims,
  mapping = aes(
    x = height_z,
    ymax = up_bound,
    ymin = lo_bound
  ),
  alpha = .1
) +
  labs(caption = "with 89% HPDI overlaid")
```

Let's learn more about link and apply. They are used to help generate predictions from the posterior


```{r}
mu <- link(m4.3)
str(mu)
View(mu)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob = 0.89)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.89)
```


Here is how to do a similar workflow with the data and code from the book.

```{r}

# data
library(rethinking)
data(WaffleDivorce)
d <- tibble(WaffleDivorce) %>%
  mutate(D = standardize(Divorce)) %>%
  mutate(M = standardize(Marriage)) %>%
  mutate(A = standardize(MedianAgeMarriage))
# model first
m5.1 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bA * A,
    a ~ dnorm(0, 0.2),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

# prior predictive simulation - accessing the priors assigned

set.seed(10)
View(prior)
prior <- extract.prior(m5.1, 1000)
mu <- link(m5.1, post = prior, data = list(A = c(-2, 2)))
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), xlab = "Median age marriage (std)", ylab = "Divorce rate (std)")
for (i in 1:50) lines(c(-2, 2), mu[i, ], col = col.alpha("black", 0.4))

# posterior predictions
A_seq <- seq(from = -3, to = 3.2, length.out = 30)
mu <- link(m5.1, data = list(A = A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- sapply(A_seq, function(x) PI(prior[["a"]] + prior[["bA"]] * x))
# plot it all
plot(D ~ A, data = d, col = rangi2)
lines(A_seq, mu.mean, lwd = 2)
shade(mu.PI, A_seq)
```

# Chapter 5

Let's try to approx the posterior of a multiple regression: chapter 5

```{r}
m5.3 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM * M + bA * A,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

# always take draws from the posterior
post <- tidy_draws(m5.3, n = 1e4) %>%
  select(a, bM, bA)
# here we are setting up for our coefficient plot, see page 22 in notes to
# understand how to read a plot... usually we plot different models together to compare
long_post <- post %>%
  pivot_longer(everything(),
    names_to = "term",
    values_to = "values"
  )

postsum <- long_post %>%
  group_by(term) %>%
  summarize(
    mean = mean(values),
    lb = quantile(values, .055),
    ub = quantile(values, .945)
  )


# here is how to plot a coeff plot
ggplot(
  postsum,
  aes(mean, term,
    xmin = lb,
    xmax = ub
  )
) +
  geom_point(size = 2) +
  geom_linerange(size = 1) +
  geom_vline(
    xintercept = 0,
    color = "gray"
  ) +
  theme(legend.position = "none") +
  xlim(-2, 2) +
  labs(
    x = "Estimate",
    y = ""
  )
```


Let's look at counter-factual plots


```{r}
m5.3_A <- quap(
  alist(
    ## A->D<-M
    D ~ dnorm(mu, sigma),
    mu <- a + bM * M + bA * A,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1),
    ## A->M
    M ~ dnorm(mu_M, sigma_M),
    mu_M <- aM + bAM * A,
    aM ~ dnorm(0, 0.2),
    bAM ~ dnorm(0, 0.5),
    sigma_M ~ dexp(1)
  ),
  data = d
)
```


Let's test out WAIC and PSIS to make sure we know how to use it. 
These are ways to measure uncertainty. So, the smaller the better.
```{r}

WAIC(m5.3)
# to see at each point, add this
WAIC(m5.3, pointwise = TRUE)
PSIS(m5.3)
# to see at each point, add this
PSIS(m5.3, pointwise = TRUE)
```



Let's try MCMC

```{r}

# install.packages("remotes")
remotes::install_github("stan-dev/cmdstanr")
install_cmdstan()
library(cmdstanr)

# data was standardized (tranformed above)
# now I am selecting what I need from the data for my model
d <- d %>%
  select(D, A, M)
View(d)

# creating inner workings of model
f <- alist(
  D ~ dnorm(mu, sigma),
  mu <- a + bM * M + bA * A,
  a ~ dnorm(0, 0.2),
  bM ~ dnorm(0, 0.5),
  bA ~ dnorm(0, 0.5),
  sigma ~ dexp(1)
)
# running with ulam
mHMC <- ulam(f,
  data = d, chains = 4, cores = 4
)
precis(mHMC)
show(mHMC)
pairs(mHMC)
traceplot(mHMC)
trankplot(mHMC)
```


```{r}
# data, start by transforming any variables needed
data(rugged)
d2 <- rugged
d2$log_gdp <- log(d2$rgdppc_2000)
dd <- d2[complete.cases(d2$rgdppc_2000), ]
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
dd$rugged_std <- dd$rugged / max(dd$rugged)
dd$cid <- ifelse(dd$cont_africa == 1, 1, 2)

# then slim data down to only what you need
dat_slim <- list(
  log_gdp_std = dd$log_gdp_std,
  rugged_std = dd$rugged_std,
  cid = as.integer(dd$cid)
)
str(dat_slim)

# lastly run model

m9.1 <- ulam(
  alist(
    log_gdp_std ~ dnorm(mu, sigma),
    mu <- a[cid] + b[cid] * (rugged_std - 0.215),
    a[cid] ~ dnorm(1, 0.1),
    b[cid] ~ dnorm(0, 0.3),
    sigma ~ dexp(1)
  ),
  data = dat_slim, chains = 4, cores = 4, control = list(adapt_delta = 0.99)
)

precis(m9.1, depth = 2)
trankplot(m9.1)
pairs(m9.1@stanfit)
```

Let's look at a model with extremely flat priors and see the errors

```{r}
y <- c(-1, 1)
set.seed(11)
m9.2 <- ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- alpha,
    alpha ~ dnorm(0, 1000),
    sigma ~ dexp(0.0001)
  ),
  data = list(y = y), chains = 3
)

precis(m9.2)

# trying option one from warning messages
m9.2 <- ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- alpha,
    alpha ~ dnorm(0, 1000),
    sigma ~ dexp(0.0001)
  ),
  data = list(y = y), chains = 3, control = list(adapt_delta = 0.99)
)
# trying option 2 from warning messages
pairs(m9.2@stanfit)
# look at how skewed these are....


# now lets look at the traceplots

traceplot(m9.2)

# they do not look anything alike... which shows us that they are not a healthy pair of chains
```

This is not what we expected, the variation is insane and our alpha should be around 0!

Notice the warning messages
Warning messages:
1: There were 47 divergent transitions after warmup. See
https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
to find out why this is a problem and how to eliminate them. 
2: Examine the pairs() plot to diagnose sampling problems


Let's work on how to plot an interaction

```{r}
library(gssr)

gss96 <- gss_get_yr(1996)
glimpse(gss96)

a <- tibble(gss96) %>%
  select(rincome, usfrustr, race) %>%
  drop_na()

a <- a %>%
  mutate(race = ifelse(race == 1, 1, 2)) %>% # create white and non-white
  mutate(logrinc = log(rincome)) %>%
  mutate(F = usfrustr)


r <- quap(
  alist(
    F ~ dnorm(mu, sigma),
    mu <- a[race] + b[race] * (logrinc),
    a[race] ~ dnorm(1, 0.1),
    b[race] ~ dnorm(0, 0.3),
    sigma ~ dexp(1)
  ),
  data = a
)
# get summary of model
precis(r, 2)
# check uncertainty
WAIC(r)
PSIS(r)

library(sjPlot)
library(ggplot2)
# this is how to plot an interaction
m2 <- lm(F ~ logrinc * race, data = a)
install.packages("sjPlot")
plot_model(m2, type = "int")
```


Also, you will need to know how to do a trankplot

```{r}
library(bayesplot)
library(aptheme)
library(rethinking)
y <- (c(-2, 2))

# now we can create our models,
# I will create one with extremely flat priors first
# to show a malfunctioning Markov chain
set.seed(1101)
p <- ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- a,
    a ~ dnorm(0, 1000),
    sigma ~ dexp(0.0001)
  ),
  data = list(y = y), chains = 3, cores = 3
)

precis(p)
# this is not good, we made our a centered on 0 and it is no where near that
# also the sigma looks insane

# let's make a trace plot

traceplot(p)

# get the stan code from the ulam model
model_stan <- stancode(p)
# fit the model in stan
model_fit <- stan(model_code = model_stan, data = list(y = y))
# then to make the trank plot use this code
mcmc_rank_overlay(model_fit) + theme_ap(family = "")
```


# Chapter 11\

```{r}
library(tidyverse)
library(gssr)
```


Messing with the data. 

```{r}

# get year 2018
d <- gss_get_yr(2018)

# manipulate for what you want
ds <- d %>%
  select(pres16, educ, sex) %>%
  drop_na() %>%
  filter(pres16 == 1 | pres16 == 2) %>%
  mutate(
    trump = ifelse(pres16 == 2, 1L, 0L),
    female = ifelse(sex == 2, 1L, 0L)
  ) %>%
  select(trump, female, educ) %>%
  mutate(male = 1 - female)
```


Using GLM. 

```{r}
# with female
m1 <- glm(trump ~ educ + female,
  data = ds,
  family = binomial(link = "logit")
)

# with male
m2 <- glm(trump ~ educ + male,
  data = ds,
  family = binomial(link = "logit")
)
# looking at it

summary(m1)
summary(m2)
```

Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  0.80867    0.30231   2.675  0.00747 ** 
educ        -0.05590    0.02037  -2.744  0.00608 ** 
female      -0.54480    0.11176  -4.875 1.09e-06 ***

The results are in log-odds. \

INTERPRETATION:\
"Each additional year of education reduces the log-odds by 0.06."\


EXAMPLE:\
What is the expected value for a male with 0 years of education?\

The model thinks a male with 0 years of education's log odds of voting for trump are 0.81.\
```{r}
0.80867 - 0.05590 * 0 - 0.54480 * 0
```


How do we get to a probability?\

```{r}
(exp(0.80867) / (1 + exp(0.80867)))
```

The model thinks that a male with 0 years of education has a 69% probability of voting for trump.

Now the man has 20 years of education. What are his log odds?

```{r}
0.80867 - 0.05590 * 20 - 0.54480 * 0
```

He has a log odds of -0.30933. 

Now, let's make it a probability. 

```{r}
(exp(-0.30933) / (1 + exp(-0.30933)))
```

The model thinks that a male with 20 years of education has a 42% chance of voting for trump. 

can do it with confidence intervals.

```{r}
broom::tidy(m1, conf.int = TRUE)
```


When education increases by one unit you can expect the log-odds of voting for trump to be between -0.096 and - 0.016. 

What's the probability of a female respondent with 12 years of education. 

```{r}
0.80867 - 0.05590 * 12 - 0.54480 * 1
```

The model thinks a female with 12 years of education has log odds of voting for trump of -0.40693

Now to probability
```{r}
(exp(-0.40693) / (1 + exp(-0.40693)))
```






Example 2: 

a man with 0 years of education and a man with 1 year of education

```{r}
# 0 years
z_log <- 0.80867 - 0.05590 * 0 - 0.54480 * 0

# now probability
z_prob <- (exp(z_log) / (1 + exp(z_log)))
# 1 year
o_log <- 0.80867 - 0.05590 * 1 - 0.54480 * 0

o_prob <- (exp(o_log) / (1 + exp(o_log)))

# difference

o_prob - z_prob
```

- 0.012044 -> - 1.2 percent


Example 3: 

a man with 19 years of education and a man with 20 year of education

```{r}
# 19 years
n_log <- 0.80867 - 0.05590 * 19 - 0.54480 * 0

# now probability
n_prob <- (exp(n_log) / (1 + exp(n_log)))

# 20 year
t_log <- 0.80867 - 0.05590 * 20 - 0.54480 * 0

t_prob <- (exp(t_log) / (1 + exp(t_log)))

# difference

t_prob - n_prob
```
- 0.0137011 -> - 1.4 percent; notice that its not consistent... the closer to 50% the bigger difference it makes (remember the S-curve)



Male vs. female voting for trump at 12 years of education

```{r}
# 12 years, female
f_log <- 0.80867 - 0.05590 * 12 - 0.54480 * 1

# now probability
f_prob <- (exp(f_log) / (1 + exp(f_log)))

# 12 years, male
m_log <- 0.80867 - 0.05590 * 12 - 0.54480 * 0

m_prob <- (exp(m_log) / (1 + exp(m_log)))

# difference

m_prob - f_prob
```
0.1347645 -> 1.34 %


Alex: "What's the effect of education?" 

(1) odds ratio

```{r}
broom::tidy(m1, exponentiate = TRUE)
```
# A tibble: 3 × 5
  term        estimate std.error statistic    p.value
  <chr>          <dbl>     <dbl>     <dbl>      <dbl>
1 (Intercept)    2.24     0.302       2.67 0.00747   
2 educ           0.946    0.0204     -2.74 0.00608   
3 female         0.580    0.112      -4.87 0.00000109


odd-ratio -> "a female respondents for voting for trump's odds are 0.580 times male respondent's odds."


(2) marginal effects
```{r}
library(marginaleffects)

marginaleffects(m1,
  newdata = ds
)


ggeffects::ggpredict(m2, terms = c("educ [0,100]", "male")) %>%
  plot()
```
