---
title: "Chapter 1 and 2"
author: "Allyson Cameron"
date: "2022-09-01"
output: html_document
---
---
title: "Chapter 1 & 2 - MD"
author: "Allyson Cameron"
date: "2022-09-01"
output: html_document
---
#Chapter 1
##Question 1  
First, I will begin by installing three packages for the examples below.

```{r}
#to install
install.packages("causact")
install.packages("dplyr")
install.packages("igraph")
```

##Question 2  
Next, we will load the packages using the `library`funtion.

```{r}
library(causact)
library(dplyr)
library(igraph)
```

Let's see what happens when we use the function `as_data_frame`!

```{r}
df <- as_data_frame(x = c(1,2,3))
```

The error says : Not a graph object. This is happening because `igraph` was the last package installed and it is using the igraph logic for the code. Let's try the code with the `dplyr` package.

```{r}
# create dataframe using dplyr package
df <- dplyr::as_data_frame(x = c(1,2,3))
glimpse(df)
```

**Why did this one work?** This code worked because we specified we wanted to use the `dplyr` package which was able to read the vector as a data frame instead of something to be graphed.

```{r}
x <- c(5,6,2,7,9,1)
n_distinct(x)
```

**Can you figure out why R called the function from 'igraph' instead of 'dplyr'?** This is happening because `igraph` was the last package installed and it is using the `igraph` logic for the code.

##Question 3
Let's find out what the `n_distinct` function does.
```{r}
?n_distinct
```
This function tells us the number,*n*, of distinct components within the vector we specify in the parentheses.

##Question 4
Now let's get a glimpse of the 'baseballData' from the package 'causeact'.
```{r}
glimpse(baseballData)
```

There are 12,145 rows and 5 columns. 
Home is a categorical/character variable representing the home team name/state. 
HomeScore is a numeric variable representing the home teams score . 

##Question 5
```{r}
baseballData[1,]
```
The  row represents case, the unit of analysis is games so this row is one game with the date, home and visitor team names/states, and their respective scores.

```{r}
baseballData[,2:3] %>% head()
```
The two columns represent variables for all of the cases in our data set (we are only showing the first 6 values because we used the `head` function). The first column is showing the name/state of the home team the second column is showing the name/state of the visitor team.

##Question 6
Let's create a data set of our top ten hockey goal scorers.

```{r}
# First, create the variables 
name <-
  c(
    "Wayne Gretzky",
    "Gordie Howe",
    "Jaromir Jagr",
    "Brett Hull",
    "Marcel Dionne",
    "Phil Esposito" ,
    "Mike Gartner",
    "Alex Ovechkin",
    "Mark Messier" ,
    "Steve Yzerman")

goals <- c(894, 801, 766, 741, 731, 717, 708, 700, 694, 692)

year_started <- c(1979, 1946, 1990, 1986, 1971, 1963, 1979, 2005, 1979, 1983)

# Now let's actually build the data frame and view it!
df <- tibble( 
  name = name, 
  goals = goals,
  year_started = year_started)

glimpse(df)
```
#Chapter 2

## Question 1
```{r}
library(tidyverse)
```

Now, let's add some data directly from the internet. 
```{r}
olympics  <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')
glimpse(olympics)
```
Now, we are going to look directly at the medal column. 
```{r}
table(olympics$medal)
```

Let's filter out for only Gold Medalist. 
```{r}
gold_medalists <- olympics %>% 
  filter(medal == "Gold")
glimpse(gold_medalists)
```

We have filtered out only gold medalist, there are 13,372 rows (cases). 
## Question 2
```{r}
library(ggplot2)
ggplot(data = gold_medalists, mapping = aes(x = year, y = age)) + geom_point(alpha = 0.1) + labs(y = "Age", x = "Year")
```
The most appropriate graph is a scatter plot because we have two numeric variables with a multitude of cases.

The age of participants has clustered around 20-40 years.

I used the alpha functionality to help view where the graph clusters, but did not use the jitter functionality because it made it more confusing on the actual ages of the participants. 

##Question 3
```{r}
us_medals <- gold_medalists %>% 
  filter(noc == "USA") %>% 
  group_by(year) %>% 
  summarise(num_medals = n())
ggplot(data=us_medals, mapping = aes(x = year, y = num_medals)) + geom_line() +labs(y= "Number of Medals", x = "Year") 

```
***What was the country's most successful year?***
1984
```{r}
# Note to self: first write what variable you're looking for, use brackets to to begin sub-setting by the condition you write next
us_medals$year[us_medals$num_medals == max(us_medals$num_medals)]
```

***As a bonus, can you guess why the line is so wiggly (technical term) towards the end?***
Since the graph points are what matter when analyzing the rate of change, when we see these wiggly lines it seems like there is a larger change between the points (maybe seasons winter & summer) during these years. For instance, the top points show that we were excelling, adn the low points show we were doing pretty terrible. 
